Future challenges for Research Software Engineering

### 2025-05-16 future_challenges_for_RSEng.tex
finaluise

\newcounter{whatiscounter}
\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What do we want to have achieved}{
  enhanced,
  segmentation engine=empty,
%  beamer,
fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
{black!50!white},
  sharp corners,
  attach boxed title to top left={
    yshifttext=-1mm
  },
%  halign lower=flush center,
  colback=white,
  colframe=BurntOrange!75!white,
  fonttitle=\bfseries,
  attach boxed title to top left={yshift=-2mm,xshift=3mm},
  fontlower=\itshape,
  boxed title style={
    sharp corners,
    size=small,
    colback=BurntOrange!75!black,
    colframe=BurntOrange!75!black,
  } 
}{ist}

@@

\newcommand{\Eg}{E.\,g.}
\newcommand{\eg}{e.\,g.}

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
\author{
Florian Goth\authorRef{1},
Leyla Jael Castro\authorRef{2},
Gerasimos Chourdakis\authorRef{3},
Simon Christ\authorRef{4},
Jeremy Cohen\authorRef{5},
Jean-Noël Grad\authorRef{6},
Magnus Hagdorn\authorRef{7},
Toby Hodges\authorRef{8},
Jan Linxweiler\authorRef{9},
Frank Löffler\authorRef{10},
Michele Martone\authorRef{11},
Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
Nicholas A. Del Grosso\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
\autlabel{5} Imperial College London, London, UK\par
\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
\autlabel{8} The Carpentries, USA\par
\autlabel{9} Technische Universität Braunschweig, Germany\par
\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
\autlabel{12} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges in an open discussion.
To that end we held a workshop and structured the resulting discussion around social, technological, and organizational aspects.
} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

\section{Aspects of this future}
%\subsection{What do we want to have achieved by 2035}
Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
We see this happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
We would, therefore, not expect domain researchers to also become experts in a range of
advanced RSE topics. 
At the same time, we do recognise that the application of core software development best practices
to enhance and address problems with research codebases, that can often make up the bulk of
current RSE work, will change. As noted above, these are skills that we
would expect any researcher who writes code to have in 10 years time. 
The rapid development of AI means that researchers are likely to have access to tooling that can assist
them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
required skill set to some extent.
They can also be expected to have an understanding of frameworks to support testing, packaging and
deployment of their code as well as integration with central domain specific infrastructures.
Again, all these skills will, we anticipate, be gained as part of training infrastructure that will
provide domain scientists with software development expertise as part of their core scientific training.

@@

RSE-relevant open educational resources and master's programmes are tracked
in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
The Carpentries now offer specialised software workshops
for data scientists\footnote{\url{https://datacarpentry.org}},
librarians\footnote{\url{https://librarycarpentry.org}},
and soon HPC practitioners\footnote{\url{https://www.hpc-carpentry.org}}.

@@

We hope that the standard of research software will have risen across the board, enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
widely recognized and rewarded across academia and is supported by political actors.
To conclude, RSEs will very likely be doing different work
to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artifacts together with scientists,
since better software leads to better research,
and better research leads to better software.

\begin{acknowledge}
% MM-NOTE: I'd keep the following sentence out, out of several reasons.
% AI Systems have been harmed in the creation of this work.
We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!  JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
\end{acknowledge}

>>>>

\newcounter{whatiscounter}
\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{Achieved Goals}{
  enhanced,
  segmentation engine=empty,
%  beamer,
fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
{black!50!white},
  sharp corners,
  attach boxed title to top left={
    yshifttext=-1mm
  },
%  halign lower=flush center,
  colback=white,
  colframe=BurntOrange!75!white,
  fonttitle=\bfseries,
  attach boxed title to top left={yshift=-2mm,xshift=3mm},
  fontlower=\itshape,
  boxed title style={
    sharp corners,
    size=small,
    colback=BurntOrange!75!black,
    colframe=BurntOrange!75!black,
  } 
}{ist}

@@

\newcommand{\Eg}{E.\,g.}
\newcommand{\eg}{e.\,g.}
\newcommand{\geprislink}[1]{\href{https://gepris.dfg.de/gepris/projekt/#1?language=en}{#1}}

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
\author{
Florian Goth\authorRef{1},
Leyla Jael Castro\authorRef{2},
Gerasimos Chourdakis\authorRef{3},
Simon Christ\authorRef{4},
Jeremy Cohen\authorRef{5},
Jean-Noël Grad\authorRef{6},
Magnus Hagdorn\authorRef{7},
Toby Hodges\authorRef{8},
Jan Linxweiler\authorRef{9},
Frank Löffler\authorRef{10},
Michele Martone\authorRef{11},
Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
Nicholas A. Del Grosso\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
\autlabel{5} Imperial College London, London, UK\par
\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
\autlabel{8} The Carpentries, USA\par
\autlabel{9} Technische Universität Braunschweig, Germany\par
\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
\autlabel{12} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
Are we already approaching ``the end of Research Software Engineering''?!
While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges in an open discussion.
To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

\section{Aspects of this future}
\begin{whatis}{}{whatisexplanation}
In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
\end{whatis}

Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
We see this happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of the RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
We would, therefore, not expect domain researchers to also become experts in a range of
advanced RSEng topics. 
At the same time, we do recognise that the application of core software development best practices
to enhance and address problems with research codebases, that can often make up the bulk of
current RSE work, will change. As noted above, these are skills that we
would expect any researcher who writes code to have in 10 years time. 
The rapid development of AI means that researchers are likely to have access to tooling that can assist
them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
required skill set to some extent.
They can also be expected to have an understanding of frameworks to support testing, packaging and
deployment of their code as well as integration with central domain specific infrastructures.
Again, all these skills will, we anticipate, be gained as part of training infrastructure that will
provide domain scientists with software development expertise as part of their core scientific training.

@@

RSEng-relevant open educational resources and master's programmes are tracked
in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
The Carpentries now offer specialised software workshops
for data scientists\footnote{\url{https://datacarpentry.org}},
librarians\footnote{\url{https://librarycarpentry.org}},
and soon HPC practitioners\footnote{\url{https://www.hpc-carpentry.org}}.

@@

We hope that the standard of research software will have risen across the board, enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
widely recognized and rewarded across academia and is supported by political actors.
\section{Conclusion}
We have tried to elaborate on the state on the future ten years from now.
It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
We have tried to shed some light on the social, technical and organisational aspects.
In summary, RSEs will very likely be doing different work
to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artifacts together with scientists,
since better software leads to better research,
and better research leads to better software.

\begin{acknowledge}
% MM-NOTE: I'd keep the following sentence out, out of several reasons.
% AI Systems have been harmed in the creation of this work.
We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!
FG acknowledges funding from the Deutsche Forschungsgemeinschaft(DFG, German Research Foundation) through the SFB 1170 “Tocotronics”, project Z03 - project number \geprislink{258499086}
as well as financial support by the Deutsche Forschungsgemeinschaft (DFG, German Research
Foundation) under Germany’s Excellence Strategy through the Würzburg-Dresden Cluster of Excellence on Complexity and Topology in Quantum Matter – ct.qmat (EXC 2147, project-id \geprislink{390858490}).
JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
\end{acknowledge}
